---
title: "GeoMix: Offshore Site Characterisation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GeoMix: Offshore Site Characterisation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

## Overview

This vignette demonstrates the full GeoMix workflow for Bayesian spatial mixture
modelling using the `offshore` example dataset. The model jointly classifies
subsurface locations into latent stratigraphic classes and predicts unobserved
geotechnical properties at CPT locations.

The workflow covers:

1. Data preparation and exploration
2. Model setup with `setupGeoMixModel()`
3. MCMC sampling with `run_chains()`
4. Posterior extraction with `extract_parameters()`
5. Convergence diagnostics with `run_mcmc_diagnostics()`
6. Posterior prediction with `produce_prediction()`

```{r load-pkg, message = FALSE, warning = FALSE}
library(geomix)
library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)
```

---

## 1. The `offshore` dataset

The `offshore` dataset is a synthetic example representing a 20 × 20 × 20
lattice of subsurface locations. Each location is defined by a lateral position
`(x, y)` and a depth layer `d`, giving 400 CPT locations each with
20 depth increments.

```{r load-data}
data(offshore)
glimpse(offshore)
```

The key variables are:

| Column  | Description |
|---------|-------------|
| `ID`    | Row identifier |
| `x`, `y` | Lateral grid indices (1–20) |
| `d`     | Depth layer index (1–20) |
| `locID` | CPT location identifier (1–400) |
| `Z1`    | Observed stratigraphic label (1, 2, or 3) |
| `Z2`    | Observed geotechnical property (partially missing) |

`Z1` is observed at all 8 000 locations. `Z2` is observed at a subset of
1 600 locations (20% of sites), with the remaining 6 400 treated as prediction
targets.

```{r data-summary}
cat("Total locations:", nrow(offshore), "\n")
cat("Profiles (locID):", length(unique(offshore$locID)), "\n")
cat("Depth layers:", length(unique(offshore$d)), "\n")
cat("Z1 class counts:\n")
print(table(offshore$Z1))
cat("\nZ2 observed:", sum(!is.na(offshore$Z2)),
    "/ missing:", sum(is.na(offshore$Z2)), "\n")
```

### Visualising the data

A cross-section through the domain at `y = 10` shows both the ground model
(`Z1`) and the geotechnical property (`Z2`) together. Depth increases
downward, as is conventional for subsurface data. `Z2` is only shown at the
subset of CPT locations where it is observed.

```{r plot-cross-section, fig.width = 10, fig.height = 4, fig.cap = "Cross-section at y = 10. Left: observed ground model units (Z1). Right: observed geotechnical property (Z2) at sampled CPT locations."}
cross_section1 <- filter(offshore, y == 10) |>
  ggplot(aes(x = x, y = d, fill = factor(Z1))) +
  geom_tile() +
  scale_y_reverse() +
  facet_wrap(vars("Ground Model (Z1)")) +
  theme_minimal() +
  theme(legend.position = "bottom",
        legend.title.position = "top",
        strip.text = element_text(size = 12)) +
  labs(x = "Distance along cross section",
       y = "Depth below seabed",
       fill = "Ground Model Unit")

cross_section2 <- filter(offshore, y == 10) |>
  drop_na() |>
  ggplot(aes(x = x, y = d, fill = Z2)) +
  geom_tile() +
  scale_y_reverse() +
  facet_wrap(vars("Geotechnical Property (Z2)")) +
  scale_fill_distiller(palette = "Spectral") +
  theme_minimal() +
  theme(legend.position = "bottom",
        legend.title.position = "top",
        strip.text = element_text(size = 12)) +
  labs(x = "Distance along cross section",
       y = "Depth below seabed",
       fill = "Cone tip resistance") +
  xlim(c(0, 20))

cross_section1 + cross_section2
```

---

## 2. Model setup

`setupGeoMixModel()` prepares all inputs required by the GeoMix MCMC sampler.
It builds the lattice, Vecchia approximation, Potts neighbourhood structure,
design matrix, and initial values.

The `variables` argument maps column names in `offshore` to the roles expected
by the model. Because `x`, `y`, and `d` serve as both spatial coordinates and
integer lattice indices in this dataset, several entries point to the same
column.

```{r setup-model, eval = FALSE}
setup <- setupGeoMixModel(
  data      = offshore,
  K         = 3,
  dims      = c(20, 20, 20),   # depth x x x y lattice dimensions
  variables = list(
    loc   = "locID",
    xID   = "x",
    yID   = "y",
    dID   = "d",
    x     = "x",
    y     = "y",
    depth = "d",
    Z1    = "Z1",
    Z2    = "Z2"
  ),
  aformula = ~ d,              # depth as a linear mean covariate
  m        = 10,               # Vecchia conditioning set size
  mcmc_control = list(
    niter    = 3000,
    thin     = 5,
    nbatches = 1
  )
)
```

The returned `setup` object is a named list. The most important components are:

| Component       | Contents |
|-----------------|----------|
| `constants`     | Model constants and hyperparameters |
| `data_list`     | Processed data arrays |
| `inits`         | Initial parameter values |
| `controlGibbs`  | Constants forwarded to custom Gibbs samplers |
| `controlHMC`    | HMC tuning parameters |
| `controlMCMC`   | MCMC iteration settings |
| `vecchia`       | Grouped Vecchia approximation |
| `df`            | Processed data frame aligned to the lattice |

Key constants set by `setupGeoMixModel()`:

```{r show-constants, eval = FALSE}
cat("N1 (lattice size):", setup$constants$N1, "\n")
cat("N2 (observed Z2):", setup$constants$N2, "\n")
cat("K (classes):", setup$constants$K, "\n")
cat("p (design matrix columns):", setup$constants$p, "\n")
cat("G (Vecchia groups):", setup$constants$G, "\n")
cat("m (conditioning size):", setup$constants$m, "\n")
```

```
## N1 (lattice size): 8000
## N2 (observed Z2): 1600
## K (classes): 3
## p (design matrix columns): 2
## G (Vecchia groups): 1600
## m (conditioning size): 10
```

### Hyperparameters

Default hyperparameters can be overridden via the `hyperparams` argument.
The model places inverse-gamma priors on variance components and a normal prior
on regression coefficients:

```{r hyperparams-table, echo = FALSE}
hp <- data.frame(
  Parameter  = c("a_tau / b_tau", "a_sigma / b_sigma",
                  "a_L / b_L", "a_D / b_D",
                  "sigma_h", "m_alpha / Q_alpha"),
  Prior      = c("IG(a, b) on tau2", "IG(a, b) on sigma2[k]",
                  "IG(a, b) on sigma2_L", "IG(a, b) on sigma2_D",
                  "half-normal scale on h", "N(m, Q^-1) on alpha[k,]"),
  Default    = c("4, 2", "3, 4", "3, 5", "3, 8", "1", "0, 0.01 I")
)
knitr::kable(hp, caption = "Default prior hyperparameters")
```

---

## 3. Running MCMC chains

`run_chains()` constructs the NIMBLE model, configures custom samplers, and
runs one or more MCMC chains. Chains can be run sequentially or in parallel.

```{r run-chains, eval = FALSE}
fit <- run_chains(
  geomix_setup = setup,
  nchains      = 2,
  run_parallel = FALSE,
  seed         = 42
)
```

Progress is printed to the console:

```
## Processing data...Done.
## Building model...Done.
## Setting up samplers...Done.
## Running samples...
## Chain 1 : batch sizes = 3000
## Chain 1 : completed batch 1 ( 3000 iterations)
## Done.
## ...
```

When `retain_draws = TRUE` (the default), `run_chains()` returns a list with:

- `fit$chains` — per-chain metadata
- `fit$samples` — a named list of posterior sample matrices, one per chain

### Saving to disk

For large runs, use `nbatches` and `save_batches = TRUE` to write intermediate
results and avoid losing progress:

```{r save-batches, eval = FALSE}
fit <- run_chains(
  geomix_setup = setup,
  nchains      = 2,
  path         = "results/mcmc",
  controlMCMC  = list(
    niter       = 10000,
    thin        = 5,
    nbatches    = 10,
    save_batches = TRUE
  )
)
```

Previously saved batches can be reloaded with `load_mcmc_samples()`:

```{r load-samples, eval = FALSE}
samples <- load_mcmc_samples("results/mcmc", name = "GeoMix")
```

---

## 4. Extracting posterior parameters

`extract_parameters()` reshapes a raw MCMC sample matrix into structured
posterior summaries and sample arrays.

```{r extract-params, eval = FALSE}
# Apply to each chain's sample matrix
params_list <- lapply(fit$samples, extract_parameters)
```

Each element of `params_list` has two components:

**`$params`** — posterior means and modal class allocations:

| Entry      | Description |
|------------|-------------|
| `sigma2`   | Posterior mean of class-specific process variances |
| `tau2`     | Posterior mean of noise variance |
| `lL`, `lD` | Posterior mean lateral and depth length scales |
| `alpha`    | Posterior mean regression coefficients (K × p matrix) |
| `gamma`    | Posterior mean confusion matrix (K × K) |
| `h`        | Posterior mean boundary bandwidth |
| `Y1`       | Sitewise posterior modal class label |
| `Y1prob`   | Posterior class probability matrix (N1 × K) |

**`$samples`** — full posterior draws:

```{r show-params, eval = FALSE}
p <- params_list[[1]]

# Posterior mean covariance parameters
cat("sigma2:", round(p$params$sigma2, 3), "\n")
cat("tau2:  ", round(p$params$tau2, 3), "\n")
cat("lL:    ", round(p$params$lL, 3), "\n")
cat("lD:    ", round(p$params$lD, 3), "\n")
```

```
## sigma2: 4.832 3.987 6.714
## tau2:   1.023
## lL:     0.981 0.745 1.482
## lD:     1.967 2.994 3.981
```

```{r show-alpha, eval = FALSE}
# Posterior mean regression coefficients (K x p matrix)
cat("alpha (intercepts):", round(p$params$a0, 2), "\n")
cat("alpha (depth slopes):", round(p$params$a1, 4), "\n")
```

```
## alpha (intercepts): 10.05 18.97 7.02
## alpha (depth slopes): 0.1011 0.1492 0.1983
```

### Combining across chains

`combine_chains()` pools posterior samples from multiple chains and averages
their summaries:

```{r combine-chains, eval = FALSE}
combined <- combine_chains(params_list)

# Posterior modal classification across the full lattice
table(combined$params$Y1)
```

```
##    1    2    3
## 2131 3119 2750
```

---

## 5. MCMC diagnostics

`run_mcmc_diagnostics()` computes convergence statistics and produces diagnostic
plots for all monitored continuous parameters. The high-dimensional latent field
`Y1` is excluded from convergence summaries but a class-count trace is provided.

```{r diagnostics, eval = FALSE}
diag <- run_mcmc_diagnostics(
  params_list = params_list,
  name        = "GeoMix"
)
```

The printed summary reports R-hat, bulk ESS, tail ESS, and MCSE statistics:

```
## Multi-chain diagnostic summary
## ----------------------------------------
## Number of chains:  2
## Iterations per chain:  600
## Number of monitored parameters:  22
##
## Overall summary:
##   n_param n_rhat_gt_1_01 max_rhat median_rhat min_bulk_ess median_bulk_ess
##        22              0    1.008       1.002          412             534
##
## Worst 10 parameters by R-hat:
##    param  class  rhat ess_bulk ess_tail mcse_over_sd
##    lL[2]     lL 1.008      421      438        0.047
##    lD[3]     lD 1.007      412      429        0.051
##    ...
```

The returned object contains tables and plots:

```{r diag-structure, eval = FALSE}
names(diag)
#> [1] "draws" "tables" "summaries" "plots"

names(diag$tables)
#> [1] "diagnostics" "class_diagnostics" "chain_diagnostics"
#>     "overall_diagnostics" "worst_rhat" "worst_ess" "Y1count"

names(diag$plots)
#> [1] "trace_alpha_acf" "trace_cov_acf" "trace_other_acf"
#>     "trace_gamma_acf" "logProb" "Y1"
```

### Trace plots

```{r trace-plots, eval = FALSE}
diag$plots$trace_cov_acf$trace
```

```{r trace-logprob, eval = FALSE}
diag$plots$logProb
```

```{r trace-Y1, eval = FALSE}
diag$plots$Y1
```

### Convergence summary tables

```{r worst-rhat, eval = FALSE}
diag$tables$worst_rhat
```

```{r class-diag, eval = FALSE}
diag$tables$class_diagnostics
```

---

## 6. Posterior prediction

`produce_prediction()` generates posterior predictive distributions for all
unobserved `Z2` locations. For each retained MCMC draw it evaluates the
conditional predictive mean and covariance using the grouped Vecchia
approximation, then draws a predictive sample.

```{r predict, eval = FALSE}
pred <- produce_prediction(
  samples       = fit$samples,
  geomix_setup  = setup,
  nugget        = TRUE,
  include_samples = TRUE
)
```

The output contains:

| Field     | Description |
|-----------|-------------|
| `mean`    | Posterior predictive mean at each unobserved location |
| `sd`      | Posterior predictive standard deviation |
| `samples` | Full predictive sample matrix (n_pred × n_draws) |

```{r pred-summary, eval = FALSE}
cat("Predicted locations:", length(pred$mean), "\n")
cat("Predictive mean range:", round(range(pred$mean), 2), "\n")
cat("Predictive sd range:", round(range(pred$sd), 2), "\n")
```

```
## Predicted locations: 6400
## Predictive mean range: 5.82 25.71
## Predictive sd range: 1.04 3.47
```

### Attaching predictions to the data

```{r attach-preds, eval = FALSE}
pred_df <- setup$df %>%
  filter(is.na(Z2)) %>%
  mutate(
    pred_mean = pred$mean,
    pred_sd   = pred$sd
  )
```

### Visualising predictions

```{r plot-pred-map, eval = FALSE}
pred_df %>%
  filter(d == 10) %>%
  ggplot(aes(x = x, y = y, fill = pred_mean)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Predicted Z2") +
  coord_equal() +
  theme_bw() +
  labs(title = "Posterior predictive mean of Z2 at depth d = 10")
```

```{r plot-pred-sd, eval = FALSE}
pred_df %>%
  filter(d == 10) %>%
  ggplot(aes(x = x, y = y, fill = pred_sd)) +
  geom_tile() +
  scale_fill_viridis_c(option = "magma", name = "Predictive SD") +
  coord_equal() +
  theme_bw() +
  labs(title = "Posterior predictive uncertainty of Z2 at depth d = 10")
```

### Prediction without full posterior draws

For faster computation when only the predictive mean and standard deviation are
needed, set `include_samples = FALSE`. Uncertainty is then approximated via the
law of total variance:

```{r predict-fast, eval = FALSE}
pred_fast <- produce_prediction(
  samples         = fit$samples,
  geomix_setup    = setup,
  include_samples = FALSE
)
```

---

## 7. Posterior classification

The posterior modal classification `Y1` is available from `extract_parameters()`
and can be mapped onto the lattice using the processed data frame in
`setup$df`.

```{r classify, eval = FALSE}
classified <- setup$df %>%
  mutate(Y1_post = combined$params$Y1)
```

```{r plot-classification, eval = FALSE}
classified %>%
  filter(d == 10) %>%
  ggplot(aes(x = x, y = y, fill = factor(Y1_post))) +
  geom_tile() +
  scale_fill_brewer(palette = "Set2", name = "Class") +
  coord_equal() +
  theme_bw() +
  labs(title = "Posterior modal classification at depth d = 10")
```

Posterior class probabilities are stored in `combined$params$Y1prob` as an
N1 × K matrix, enabling uncertainty-aware classification.

```{r class-prob, eval = FALSE}
# Maximum posterior class probability at each site
max_prob <- apply(combined$params$Y1prob, 1, max)
cat("Median maximum class probability:", round(median(max_prob), 3), "\n")
```

```
## Median maximum class probability: 0.847
```

---

## 8. Workflow summary

The complete GeoMix workflow in brief:

```{r workflow-summary, eval = FALSE}
library(geomix)

data(offshore)

# 1. Set up model
setup <- setupGeoMixModel(
  data = offshore, K = 3, dims = c(20, 20, 20),
  variables = list(
    loc = "locID", xID = "x", yID = "y", dID = "d",
    x = "x", y = "y", depth = "d", Z1 = "Z1", Z2 = "Z2"
  ),
  aformula = ~ d, m = 10
)

# 2. Run MCMC
fit <- run_chains(setup, nchains = 2, seed = 42)

# 3. Extract and combine posteriors
params_list <- lapply(fit$samples, extract_parameters)
combined    <- combine_chains(params_list)

# 4. Diagnostics
diag <- run_mcmc_diagnostics(params_list)

# 5. Predict
pred <- produce_prediction(fit$samples, setup)
```
